\Large \bf Cyclone: Replication Middleware for NVM Applications

### 2016-09-22 paper/paper.tex
Updates to paper.

\section{Introduction}
Memory technology in the datacenter is due to undergo a paradigm shift with the
increased use of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as 3D
XPoint. In anticipation, libraries that allow programmers to build durable data
structures on non-volatile heaps have become available - an example of a such a
library that we have used in this paper is Intel's non volatile memory library
(NVML). A missing piece however is a component to make such durable data
structures available using replication across a commodity network such as
ethernet. Existing solutions that offer good performance are biased towards pure
transactions across RDMA with a dependence on external services and application
specific code for fault recovery [FARM, etc]. This is a problem for programmers
using NVML to construct durable and available datastructures where one cannot
apriori guarantee the availability of RDMA in the target environment especially
when running on externally hosted environments such as the cloud.

Cyclone is a prototype replication middleware that aims to demonstrate that one
can add fault tolerance transparently to a specific class of client server
applications written on top of libraries such as NVML. We specifically target
key-value maps in this paper and in that context make two key
contributions. First, we show that a simple API allows server state to be
replicated without the programmer needing to worry about fault recovery. This
allows one to start by building a concurrent durable data structure using
libraries such as NVML and add availability as a later step using Cyclone. The
second contribution we make is to show that one can achieve reasonably good
replication performance using commodity ethernet via innovations in the
networking stack and by scaling replication with the application. This result
should be a useful design point for developers looking to deploy applications on
non-volatile heaps in commodity environments such as public clouds.

\section{Single Node Programming Model}
Cyclone starts with the assumption that the programmer has already built a
concurrent durable application on a non-volatile heap. A number of libraries and
filesystems make this possible today [DAX, NVML, others ?]. As a starting point,
we support applications built on top of Intel's Non Volatile Memory Library
(NVML) [nvml.io]. We describe NVML's programming model in this section to
provide adequate background for the rest of this paper. The primary feature of
interest is the ability to write code that is crash consistent. In other words
code that makes changes to the non-volatile heap is enclosed in a crash
consistent transaction such that either all of the transaction executes or none
of it executes in the face of power failures. This is critical to maintaining
consistency as the data structure, being durable, must survive across power
failures. To illustrate this idea, Figure~\ref{fig:example} shows how one might
build a persistent linked list using NVML - the key idea is to enclose the
entire process of updating the linked list within a crash consistent
transaction. This ensure that the linked list is not corrupted due to a power
failure. A second point worth noting is that NVM libraries such as NVML usually
provide a way to wrap pointers to non-volatile memory (using {\tt TOID} macros)
in order that such memory may be mapped to different locations in virtual memory
across restarts without impact to the accessing application.

@@

It is easy to turn the linked list example in NVML into a concurrent linked list
by simply adding a lock to it. Such a lock can be maintined in volatile DRAM if
so desired for performance. More sophisticated fine-grained or optimistic
concurrency control mechanisms for NVM are also possible [mnemosyne].

@@

Cyclone provides strongly consistent replication - every machine in the quorum
of replicas maintains a log of RPC calls. Cyclone treats the RPC call itself as
a variable sized opaque blob of data, leaving the application free to choose how
it wishes to marshall call related details and argument. All machines agree on
the committed sequence of RPC calls in the log. We achieve this by running an
instance of the RAFT [raft] consensus protocol to keep the log of RPC calls on
different machines in synchronization. Specially flagged `read-only` RPC calls
are executed on the leader replica and are not replicated.

@@

\subsection{Network Stack}
Cyclone is designed around DPDK [dpdk] a high performance network stack. DPDK
provides convenient access to packets arriving at an ethernet NIC by moving them
into packet buffers with the contained cache lines also directly placed into the
last level cache of the CPU during DMA [ddio].

@@

We exploit the fact that DPDKs packet buffers allow prepending data to a packet
without physically moving the data in the packet buffer. Fig
[figures/packets.bmp] shows how the layout of a packet buffer changes at the
leader during replication. The received packets contains only the RPC call from
the client. We prepend a RAFT related header (containing the previous log index,
previous log term, leader's term and leader's commit point) to packet buffer
itself. To this RAFT header we then prepend an IPv4 header. We set IP addresses
in the header to use 5-tuple steering in order to place the packet in the right
queue at the receiving NIC. We do not use different ip addresses for different
replicas, rather the ip addresses are only set in conjunction with NIC
programming to enable steering of the received packets to the correct NIC queue.
This composite packet buffer is then sent \emph{unchanged} to every replica. In
order to properly route the packet through switches, we allocate a small buffer
containing an ethernet header for each follower replica and \emph{chain} it in
front of the prepared packet - ensuring we do not need to change the contents of
the packet itself for every replica.

Although the common case path in RAFT is straightforward, it still adds many
hundreds of cycles for each log entry to be processed. This adds up to a few
microseconds, a significant amount in the face of sub 10 us network latency
between nodes with DPDK. We amortize this overhead when operating under load by
batching multiple RPC calls together and treating the batch as a single RAFT log
entry.  DPDK by itself can efficiently batch incoming packets under load -
allowing the user to easily receive a burst (currently limited to 32) of packets
rather than a single one for a polling call. DPDK further allows chaining packet
buffers together and treating the chain as a single packet. This is convenient
for our zero copy approach and we chain packet mbufs together to form a single
packet as shown in Fig [figures/chain.bmp]. Since the entire chain is sent as a
single packet when being replicated, the length of a chain is limited by the
size of the coalesced packet, that must fit into an ethernet jumbo frame of
approximately 9KB.

@@

\begin{figure*}
\begin{tabular}{cc}
\begin{minipage}{0.5\textwidth}
  \includegraphics[width=\textwidth,height=6cm]{results/replication_mops_0.pdf}
\end{minipage}&
\begin{minipage}{0.5\textwidth}
  \includegraphics[width=\textwidth,height=6cm]{results/replication_gbps_0.pdf}
\end{minipage}\\
\begin{minipage}{0.5\textwidth}
  \includegraphics[width=\textwidth,height=6cm]{results/replication_mops_512.pdf}
\end{minipage}&
\begin{minipage}{0.5\textwidth}
  \includegraphics[width=\textwidth,height=6cm]{results/replication_gbps_512.pdf}
\end{minipage}
\end{tabular}
\caption{Pure Replication}
\label{fig:pure_rep}
\end{figure*}

@@

\begin{figure}
  \includegraphics[width=0.5\textwidth,height=6cm]{results/cc_mops.pdf}
  \caption{Crash Consistency + Replication}
  \label{fig:cc_rep}
\end{figure}

\begin{figure}
  \includegraphics[width=0.5\textwidth,height=6cm]{results/app_mops.pdf}
  \caption{Crash Consistent Replicated Hash Table}
  \label{fig:app_rep}
\end{figure}

\section{Conclusion}
Cyclone allows developers to add availability to applications built on top of
libraries such as NVML and achieve good performance on commodity networking
hardware. A key question is whether the ideas in Cyclone extend to more than
just key-value maps - perhaps to application such as filesystems. We believe
that the requirement for scaling across multiple replication quorums is
can be satisfied by applications where the scalable commutativity rule [scalable
commutativity] applies to common case primitives. Although our intitial release
of Cyclone will be targeted towards key value style applications, our future
research is targeted towards investigating this hypothesis.
\end{document}

>>>>

\section{Introduction}
Memory technology in the datacenter is due to undergo a paradigm shift with the
increased use of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as 3D
XPoint. In anticipation, libraries that allow programmers to build durable data
structures on non-volatile heaps have become available - an example of a such a
library that we have used in this paper is Intel's non volatile memory library
(NVML). A missing piece however is a component to make such durable data
structures available using replication across a commodity network such as
ethernet. Existing solutions that offer good performance are biased towards pure
transactions across RDMA with a dependence on external services (such as
zookeeper~\cite{zookeeper}) and application specific code for fault recovery~\cite{farm,
  htm}. This is a problem for programmers using NVML to construct durable and
available datastructures where one cannot apriori guarantee the availability of
RDMA in the target environment especially when running on externally hosted
environments such as the cloud.

Cyclone is a prototype replication middleware that aims to demonstrate that one
can add fault tolerance transparently to a specific class of client server
applications written on top of libraries such as NVML. We specifically target
key-value maps in this paper and in that context make two key
contributions. First, we show that a simple API allows server state to be
replicated without the programmer needing to worry about fault recovery. This
allows one to start by building a concurrent durable data structure using
libraries such as NVML and add availability as a later step using Cyclone. The
second contribution we make is to show that one can achieve reasonably good
replication performance using commodity ethernet via innovations in the
networking stack and by scaling replication with the application. This result
should be a useful design point for developers looking to deploy applications on
non-volatile heaps in commodity environments such as public clouds as well as
library builders that aim to support them.

\section{Single Node Programming Model}
Cyclone starts with the assumption that the programmer has already built a
concurrent durable application on a non-volatile heap. A number of libraries and
filesystems make this possible today~\cite{nvml, dax}. As a starting point, we
support applications built on top of Intel's Non Volatile Memory Library
(NVML~\cite{nvml}). We describe NVML's programming model in this section to
provide adequate background for the rest of this paper. The primary feature of
interest is the ability to write code that is crash consistent. In other words
code that makes changes to the non-volatile heap is enclosed in a crash
consistent transaction such that either all of the transaction executes or none
of it executes in the face of power failures. This is critical to maintaining
consistency as the data structure, being durable, must survive across power
failures. To illustrate this idea, Figure~\ref{fig:example} shows how one might
build a persistent linked list using NVML - the key idea is to enclose the
entire process of updating the linked list within a crash consistent
transaction. This ensure that the linked list is not corrupted due to a power
failure. A second point worth noting is that NVM libraries such as NVML usually
provide a way to wrap pointers to non-volatile memory (using {\tt TOID} macros)
in order that such memory may be mapped to different locations in virtual memory
across restarts without impact to the accessing application.

@@

It is easy to turn the linked list example in NVML into a concurrent linked list
by simply adding a lock to it. Such a lock can be maintined in volatile DRAM if
so desired for performance. More sophisticated fine-grained or optimistic
concurrency control mechanisms for NVM are also possible~\cite{mnemosyne}.

@@

Cyclone provides strongly consistent replication - every machine in the quorum
of replicas maintains a log of RPC calls. Cyclone treats the RPC call itself as
a variable sized opaque blob of data, leaving the application free to choose how
it wishes to marshall call related details and argument. All machines agree on
the committed sequence of RPC calls in the log. We achieve this by running an
instance of the RAFT~\cite{raft} consensus protocol to keep the log of RPC calls on
different machines in synchronization. Specially flagged `read-only` RPC calls
are executed on the leader replica and are not replicated.

@@

\subsection{Network Stack}
Cyclone is designed around DPDK~\cite{dpdk} a high performance network stack. DPDK
provides convenient access to packets arriving at an ethernet NIC by moving them
into packet buffers with the contained cache lines also directly placed into the
last level cache of the CPU during DMA~\cite{ddio}.

@@

We exploit the fact that DPDKs packet buffers allow prepending data to a packet
without physically moving the data in the packet buffer. The received packets
contains only the RPC call from the client. We prepend a RAFT related header
(containing the previous log index, previous log term, leader's term and
leader's commit point) to packet buffer itself. To this RAFT header we then
prepend an IPv4 header. We set IP addresses in the header to use 5-tuple
steering in order to place the packet in the right queue at the receiving
NIC. We do not use different ip addresses for different replicas, rather the ip
addresses are only set in conjunction with NIC programming to enable steering of
the received packets to the correct NIC queue.  This composite packet buffer is
then sent \emph{unchanged} to every replica. In order to properly route the
packet through switches, we allocate a small buffer containing an ethernet
header for each follower replica and \emph{chain} it in front of the prepared
packet - ensuring we do not need to change the contents of the packet itself for
every replica.

Although the common case path in RAFT is straightforward, it still adds many
hundreds of cycles for each log entry to be processed. This adds up to a few
microseconds, a significant amount in the face of sub 10 us network latency
between nodes with DPDK. We amortize this overhead when operating under load by
batching multiple RPC calls together and treating the batch as a single RAFT log
entry.  DPDK by itself can efficiently batch incoming packets under load -
allowing the user to easily receive a burst (currently limited to 32) of packets
rather than a single one for a polling call. DPDK further allows chaining packet
buffers together and treating the chain as a single packet. This is convenient
for our zero copy approach and we chain packet mbufs together to form a single
packet. Since the entire chain is sent as a single packet when being replicated,
the length of a chain is limited by the size of the coalesced packet, that must
fit into an ethernet jumbo frame of approximately 9KB.

@@

\begin{figure}
\begin{tabular}{cc}
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_0.pdf}
\end{minipage}&
\begin{minipage}{0.25\textwidth}
  \hspace{-0.2in}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_0.pdf}
\end{minipage}\\
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_512.pdf}
\end{minipage}&
\begin{minipage}{0.25\textwidth}
  \hspace{-0.2in}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_512.pdf}
\end{minipage}
\end{tabular}
\caption{Pure Replication}
\label{fig:pure_rep}
\end{figure}

@@

\begin{figure}
\begin{tabular}{cc}
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/cc_mops.pdf}
  \caption{Crash Consistency + Replication}
  \label{fig:cc_rep}
\end{minipage} &
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/app_mops.pdf}
  \caption{Crash Consistent Replicated Hash Table}
  \label{fig:app_rep}
\end{minipage}
\end{tabular}
\end{figure}

\section{Conclusion}
Cyclone allows developers to add availability to applications built on top of
libraries such as NVML and achieve good performance on commodity networking
hardware. A key question is whether the ideas in Cyclone extend to more than
just key-value maps - perhaps to application such as filesystems ? We believe
that the requirement for scaling across multiple replication quorums can be
satisfied by applications where the scalable commutativity
rule~\cite{scalable_commutativity} applies to common case primitives. Although
our intitial release of Cyclone will be targeted towards key value style
applications, our future research is targeted towards investigating this
hypothesis.
\newcommand\myurl[2]{\url{#1}}
\bibliographystyle{acm}
\bibliography{paper}

\end{document}
